%%% The main file. It contains definitions of basic parameters and includes all other parts.

%% Settings for single-side (simplex) printing
% Margins: left 40mm, right 25mm, top and bottom 25mm
% (but beware, LaTeX adds 1in implicitly)
%\documentclass[12pt,a4paper]{report}
%\setlength\textwidth{145mm}
%\setlength\textheight{247mm}
%\setlength\oddsidemargin{15mm}
%\setlength\evensidemargin{15mm}
%\setlength\topmargin{0mm}
%\setlength\headsep{0mm}
%\setlength\headheight{0mm}
% \openright makes the following text appear on a right-hand page
%\let\openright=\clearpage

%% Settings for two-sided (duplex) printing
\documentclass[12pt,a4paper,twoside,openright]{report}
% \setlength\textwidth{145mm}
% \setlength\textheight{247mm}
% \setlength\oddsidemargin{14.2mm}
% \setlength\evensidemargin{0mm}
% \setlength\topmargin{0mm}
% \setlength\headsep{0mm}
% \setlength\headheight{0mm}
\let\openright=\cleardoublepage

%% Generate PDF/A-2u
\usepackage[a-2u]{pdfx}

%% Character encoding: usually latin2, cp1250 or utf8:
\usepackage[utf8]{inputenc}

%% Prefer Latin Modern fonts
\usepackage[mono=false]{libertinus}

%% Further useful packages (included in most LaTeX distributions)
\usepackage{amsmath}        % extensions for typesetting of math
\usepackage{amsfonts}       % math fonts
\usepackage{amsthm}         % theorems, definitions, etc.
\usepackage{bbding}         % various symbols (squares, asterisks, scissors, ...)
\usepackage{bm}             % boldface symbols (\bm)
\usepackage{graphicx}       % embedding of pictures
\usepackage{fancyvrb}       % improved verbatim environment
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[natbib,backend=bibtex]{biblatex}         % citation style AUTHOR (YEAR), or AUTHOR [NUMBER]
\addbibresource{bibliography}
\usepackage[nottoc]{tocbibind} % makes sure that bibliography and the lists
			    % of figures/tables are included in the table
			    % of contents
\usepackage{dcolumn}        % improved alignment of table columns
\usepackage{booktabs}       % improved horizontal lines in tables
\usepackage{paralist}       % improved enumerate and itemize
\usepackage{xcolor}         % typesetting in color

\usepackage[textsize=tiny, linecolor=red!50!black, backgroundcolor=yellow!30]{todonotes}
\setlength{\marginparwidth}{6em}
\setuptodonotes{fancyline}
\newcommand{\xxx}[1]{\textcolor{red}{#1}}

%%% Basic information on the thesis

% Thesis title in English (exactly as in the formal assignment)
\def\ThesisTitle{Mahalanobis based hierarchical clustering accelerated on GPU}

% Author of the thesis
\def\ThesisAuthor{Bc. Adam Šmelko}

% Year when the thesis is submitted
\def\YearSubmitted{2020}

% Name of the department or institute, where the work was officially assigned
% (according to the Organizational Structure of MFF UK in English,
% or a full name of a department outside MFF)
\def\Department{Department of Software Engineering}

% Is it a department (katedra), or an institute (ústav)?
\def\DeptType{Department}

% Thesis supervisor: name, surname and titles
\def\Supervisor{RNDr. Miroslav Kratochvil}

% Supervisor's department (again according to Organizational structure of MFF)
\def\SupervisorsDepartment{Department of Software Engineering}

% Study programme and specialization
\def\StudyProgramme{Computer Science}
\def\StudyBranch{Software and Data Engineering}

% An optional dedication: you can thank whomever you wish (your supervisor,
% consultant, a person who lent the software, etc.)
\def\Dedication{%
Dedication.
}

% Abstract (recommended length around 80-200 words; this is not a copy of your thesis assignment!)
\def\Abstract{%
During a clustering of a flow-cytometry data, the Mahalanobis distance function is used to form elliptical clusters naturally. However, reasonably computable dataset sizes are restricted to a factor of 100K points due to~the~polynomial time and space complexity of the current algorithm. In this thesis, we researched possible ways of implementing clustering algorithms and proposed an~implementation of Mahalanobis-based hierarchical clustering analysis accelerated on a single GPU. Using CUDA platform we achieved the performance increase of up to 5000-times on single-point datasets and 40-times on apriori datasets compared to a serial CPU counterpart. With this contribution, the size of flow cytometry datasets can increase to millions of points.
}

% 3 to 5 keywords (recommended), each enclosed in curly braces
\def\Keywords{%
{clustering} {high-dimensional data} {GPU}
}

%% The hyperref package for clickable links in PDF and also for storing
%% metadata to PDF (including the table of contents).
%% Most settings are pre-set by the pdfx package.
\hypersetup{unicode}
\hypersetup{breaklinks=true}

% Definitions of macros (see description inside)
\include{macros}

% Title page and various mandatory informational pages
\begin{document}
\include{title}

%%% A page with automatically generated table of contents of the master thesis

\tableofcontents

%%% Each chapter is kept in a separate file
\include{preface}

\include{clustering}
\include{implementation}
\include{results}

\include{epilog}

%%% Bibliography
\include{bibliography}

%%% Figures used in the thesis (consider if this is needed)
%\listoffigures

%%% Tables used in the thesis (consider if this is needed)
%%% In mathematical theses, it could be better to move the list of tables to the beginning of the thesis.
%\listoftables

%%% Abbreviations used in the thesis, if any, including their explanation
%%% In mathematical theses, it could be better to move the list of abbreviations to the beginning of the thesis.
%\chapwithtoc{List of Abbreviations}

%%% Attachments to the master thesis, if any. Each attachment must be
%%% referred to at least once from the text of the thesis. Attachments
%%% are numbered.
%%%
%%% The printed version should preferably contain attachments, which can be
%%% read (additional tables and charts, supplementary text, examples of
%%% program output, etc.). The electronic version is more suited for attachments
%%% which will likely be used in an electronic form rather than read (program
%%% source code, data files, interactive charts, etc.). Electronic attachments
%%% should be uploaded to SIS and optionally also included in the thesis on a~CD/DVD.
%%% Allowed file formats are specified in provision of the rector no. 72/2017.
\appendix
\chapter{User guide}

\section{Build guide}

The Mahalanobis-based hierarchical clustering project was developed with the CMake build tool. To build the executable, use CMake configure and build commands in a build directory. Then, the directory \texttt{para} will contain \texttt{gmhclust} executable. The only dependency is the CUDA compiler (\texttt{nvcc}). The executable should be portable to all platforms supporting \texttt{nvcc}; it was successfully tested on Ubuntu 18.04 and Windows 10. See the following steps:

\begin{lstlisting}
cd gmhc
mkdir build && cd build
cmake ..
cmake --build .
ls para/gmhclust
\end{lstlisting}

\section{Running the program}

The \texttt{gmhclust} executable has three command line parameters:
\begin{enumerate}
	\item \emph{Dataset file path} -- The mandatory parameter with a path to a dataset file. The file is binary and has structure as follows:
	\begin{enumerate}
		\item 4B unsigned integer $D$ -- point \emph{dimension}
		\item 4B unsigned integer $N$ -- \emph{number} of points
		\item $N\cdot D$ 4B floats -- $N$ single-precision $D$-dimensional points stored one after another
	\end{enumerate}
	\item \emph{Mahalanobis threshold} -- An absolute positive number that states the Mahalanobis threshold. It is the mandatory parameter.
	\item \emph{Apriori assignments file path} -- An optional path to an apriori assignments file --- a file with space separated 4B unsigned integers (assignment numbers). The number of integers is the same as the number of points in the dataset; it sequentially assigns each point in the dataset file an assignment number. Then simply, if the $i$-th and the $j$-th assignment numbers are equal, then the $i$-th and $j$-th points are assigned the same apriori cluster. 
\end{enumerate}

The executable writes the clustering process to the standard output. Each line contains an ID pair of merged clusters with their merge distance as well.

\chapter{Enclosed CD}

The enclosed CD contais four folders:

\begin{itemize}
	\item \texttt{gmhc} contains the source code of the implementation.
	\item \texttt{data} contains dataset files used in experiments.
	\item \texttt{benchmark} contains scripts and a guide how to reproduce some of the performed experiments.
	\item \texttt{docs} contains the programmer documentation of the implementation.
\end{itemize} 


\openright
\end{document}
